---
title: "Functional Analysis Practical"
output: html_document
---
## Instal & Load Required Pacakges
### CRAN packages
Install CRAN packages

```{r}
cran.packages <- c("tidyverse",
                   "ggrepel",
                   "kableExtra",
                   "ggupset")

cran.load <- function(pkg){
        new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
        if (length(new.pkg)){
          install.packages(new.pkg, dependencies = TRUE)
          }
        sapply(pkg, require, character.only = TRUE)
}
cran.load(cran.packages)
```

### Bioconductor Packages
Install Biocondustor packages

```{r}
bioconductor.packages <- c("biomaRt",
                           "org.Hs.eg.db",
                           "clusterProfiler",
                           "enrichplot",
                           "pathview")

bioconductor.load <- function(pkg){
        new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
        if (length(new.pkg)){
          if (!requireNamespace("BiocManager", quietly = TRUE))
            install.packages("BiocManager")
            BiocManager::install(new.pkg)
        }
        sapply(pkg, require, character.only = TRUE)
}
bioconductor.load(bioconductor.packages)
```

## Load Save Data
Load the example results file, the data is in the ./data/ directory and called "Treatment_vs_Control.results.tsv". Note the use of the relative path. The "relative" path to the file is relative to the current working directory. Use the variable name "de_results". Complete the code chunk below.
If you don't have the results file then you can get a copy from the data directory in this Session.

```{r read in previous days results file}
de_results <- 
```

## Annotations
Using the example from the lecture slides get annotations from the Ensembl human database at Biomart. Replace the "..." with your code.
```{r}
database <- "hsapiens_gene_ensembl"
mart <- "genes"
filt <- "ensembl_gene_id"

ensembl <- useEnsembl(...)

att <- c("ensembl_gene_id","external_gene_name","chromosome_name","start_position","end_position","gene_biotype","entrezgene_id")

annotations <- getBM(...) %>% 
  distinct(ensembl_gene_id, .keep_all = TRUE)
```

Merge the annotations with the DESeq2 results
```{r}
de_results <- de_results %>%
  left_join(...)%>% # merge annotations by ensembl gene id
  arrange(...) # Sort the dataframe by the p adjusted value ascending
```

Save the annotated results to a file in the data directory, call the file "Treatment_vs_Control.annotated.results.tsv"
```{r}

```

## Results Directory
Create a results directory for the ClusterProfiler results files, I would call it "ClusterProfiler", but you can call it whatever you would like. 
HINT: Code for this is in the lecture slides
```{r}
results_directory <- 
```

## Filtering
Filter the results table for genes that have a p adjusted value of 0.1 and a log fold change of 1.
NOTE: The use of the absolute value for the log2 fold change. Complete the code chunk below

```{r}
pvalueCutoff <- ... # P value
qvalueCutoff <- ... # Adjusted P value
foldchange <- ... # Fold change, usually fold change is log2
```


```{r}
filtered_de_results <- de_results %>%
  filter(...) %>% # Filter on p adjusted and the absolute values for the log fold change
  drop_na() # Remove any NA values
```

## Gene Ontology Analysis
### Over representation Analysis
Extract a list of entrez ids for the differentially expressed genes, de_results. Use the pull() command from dplyr.  
Complete the code chunk below
```{r}
de_genes <- filtered_de_results %>%
  ...
```


Using the example in the lecture slides perform the over representation analysis
```{r}
go_ora <- enrichGO(gene = ...,
                   OrgDb = ..., # Organism DB
                   ont = ..., # Can choose "BP","CC", "MF" or "ALL"
                   pAdjustMethod = ..., # Can choose "holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none"
                   pvalueCutoff = ...,
                   qvalueCutoff = ...,
                   readable = TRUE)
```

### Gene Set enrichment Analysis

Input is a named vector of log fold changes with entrez ids as the names, the list should be sorted on the log fold changes.  
HINT: Example can be fould in the lecture slides.
```{r}
gsea_gene_list <- ... # Get the log fold values in a character vector
names(gsea_gene_list) <- ... # Add the entrez gene ids as the names for the log fold changes
gsea_gene_list <- ... # Sort the character vector bym logo fold change, for lowest to highest†
```

Using the example in the lecture slides perform the g ene set enrichment analysis
```{r}
#| echo: true
#| eval: true
go_gsea <- gseGO(gene = ..., # Add the gene list here
                 OrgDb = ..., # Organism DB
                 ont = ..., # Can choose BP,CC, MF or ALL
                 pAdjustMethod = ..., # Can choose "holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none"
                 minGSSize    = 100,
                 maxGSSize    = 500,
                 pvalueCutoff = ...)
```


Now save the Gene Ontology results to a file.

```{r}
write_delim(as.data.frame(go_ora), 
            file = file.path(results_directory, "go_ora_enriched.tsv"), 
            delim = "\t")
```


Now using the dotplot() command visualise the GO analysis results, You can limit the number of results using the parameter showCategory=10.

```{r}

```

These visualisations are ggplot2 objects and can be saved using ggsave.

## Pathway Analysis
Over representation analysis
Use the over represented gene list, de_genes, from above
```{r}
kegg_ora <- enrichKEGG(gene = ...,
                       organism = ..., # Organism DB
                       pvalueCutoff = 0.01)
```

Save the pathway analysis to file, name the file "ora_pathway.tsv"

```{r}

```

## Pathway Analysis
Gene set enrichment analysis using the named gene set list, gsea_gene_list, from above

```{r}
kegg_gsea <- gseKEGG(geneList = ...,
                     organism = ..., # Organism DB
                     minGSSize = 10,
                     pvalueCutoff = 0.05)
```

Save the pathway analysis to file, name the file "gsea_pathway.tsv"

```{r}

```


## Pathway Analysis Visualisation
Using the function function to retrieve the KEG pathways.

Get a list of KEGG path way ids, kegg_pathway_ids_list, from the kegg_gsea results and pass them thpo the function

```{r}
# Pathway visualisation function
kegg_pathview <- function(kegg_pathway_id){ 
  pathview(gene.data = gsea_gene_list, # Named list of fold changes
        pathway.id = kegg_pathway_id, # KEGG pathway id
        species = "hsa", # KEG species id
        gene.idtype = "KEGG",
        kegg.native = TRUE) 
}

# List of KEGG pathway ids
kegg_pathway_ids_list <- kegg_gsea %>%
  ...

# Iterate over the list of pathway ids and apply the data to the pathway
lapply(kegg_pathway_ids_list, kegg_pathview)
```


You can use the code below to tidy up the KEGG pathway files
```{r}
results.files <- list.files(path = ".", pattern = "pathview.png") %>% # list all pathview files
  file.copy(results_directory) # Copy pathway files to results directory
# Delete results files
remove_files <- list.files(path = ".", pattern = "hsa") %>% # List all files dowloaded from KEGG
  file.remove() # Delete the files
```

